Viscous Transport in Eroding Porous Media

### 2019-08-05 docs/transport/erosion_transport.tex
modify figures caption of transport paper
small changes

Towards the end of the erosion simulation, the porosity is low, and we
expect that the ensemble moments behave similarly to transport in a
pipe.  In a grainless geometry, the trajectory originating at
$(x_0,y_0)$ is
\begin{align}
  s_1(t) &= (1-y_0^2)t + x_0, \\
  s_2(t) &= y_0.
\end{align}
Averaging over $y_0 \in [-1,1]$, the first and second ensemble moments
are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry quantifies the average amount of turning
made by a streamline.  The choice of how to average is not universal,
and Duda et al.~\cite{dud-koz-mat2011} summarize several options for
averaging.  One particular choice, which we shall adopt, allows us to
compute the tortuosity in terms of trajectories, or by using an area
integral over the fluid geometry.  Another advantage of this choice is
that it is not affected by regions of recirculation.  Mathematically,
the definition of tortuosity that we use is
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is an cross-section perpendicular to the $x$ axis with length
$d$, $\lambda(y)$ is the length of the streamline that is initialized
at the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of
the velocity at the $y$-coordinate along $S$.
Following~\cite{dud-koz-mat2011}, the trajectories are formed until they
cross a second cross-section that is parallel to the initial
cross-section.  In our work, we always initialize the trajectories at
$x=-1$ and stop them when they reach $x=1$.

Under the assumptions that the flow is incompressible and not reentrant,
the tortuosity can be equivalently measured as~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy}.
  \label{eqn:tortuosity2}
\end{align}
The flow is reentrant if all streamlines connect the inlet to the
outlet.  Recirculation zones are possible in viscous fluids, but they
only constitute a small fraction of the total porous media and tend to
have very small velocities.  Therefore, the tortuosity defined in
equation~\eqref{eqn:tortuosity2} is an upper bound for the definition in
equation~\eqref{eqn:tortuosity1}, but for the geometries we consider, we
will show that their difference is negligible.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories, we use this definition for the
tortuosity.

The first method to compute numerically compute the tortuosity uses
$N_p$ trajectories that are initially at $(1,y_i)$, $i=1,\ldots,N_p$,
and stopped when their $x$ coordinate is 1.  We choose $y_i$ so that the
trajectories are initially uniformly spaced in $(-1,1)$.  Having
computed these trajectories, the tortuosity is approximated as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_x(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_x(y_i) \Delta y_i}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.  Alternatively, to compute the tortuosity in terms of
the Eulerian velocity field, we first compute the velocity at $(x_i,y_j)
= (-1 + i\Delta x, -1 + j\Delta y)$, $i,j=1,\ldots,N$, where $\Delta x =
\Delta y = 2/N$, and the velocity at points inside an eroding body are
assigned a value of 0.  Then, the tortuosity is approximated as
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(x_i,y_j)\|\Delta x \Delta y}
      {\sum\limits_{i,j=1}^N u(x_i,y_j) \Delta x \Delta y},
\end{align}
where $\uu = (u,v)$.

There have been efforts to relate the tortuosity to the
porosity~\cite{matyka2008tortuosity}, but these results can not be
universal since the porosity which depends on only a single number, the
tortuosity is affected by the distribution of pores within the porous
network.  In addition to the hydraulic tortuosity, a local tortuosity
can be used to analyze the local behavior of a streamline near a
body~\cite{matyka2008tortuosity}.  The local tortuosity of a streamline
initialized at $y$ along $S$ is
\begin{align}
  \tau(y) = \frac{\lambda(y)}{d}.
  \label{eqn:localTort}
\end{align}
\todo[inline]{Need to flesh this out more}

>>>>

Towards the end of the erosion simulation, the porosity is low, and we
expect that the ensemble moments behave similarly to transport in a
pipe.  In a grainless geometry, the trajectory originating at
$(s_1(0),s_2(0))$ is
\begin{align}
  s_1(t) &= (1-s_2(0)^2)t + s_1(0), \\
  s_2(t) &= s_2(0).
\end{align}
Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
moments are
\begin{align}
  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity of a geometry quantifies the average amount of turning
made by a streamline.  The choice of how to average is not universal,
and Duda et al.~\cite{dud-koz-mat2011} summarize several options for
averaging.  One particular choice, which we shall adopt, allows us to
compute the tortuosity in terms of trajectories, or by using an area
integral over the fluid geometry.  Another advantage of this choice is
that it is not affected by regions of recirculation.  Mathematically,
the definition of tortuosity that we use is
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_x(y)\lambda(y)\,dy}
  {\displaystyle\int_{S}u_x(y)\,dy},
  \label{eqn:tortuosity1}
\end{align}
where $S$ is an cross-section perpendicular to the $x$ axis with length
$d$, $\lambda(y)$ is the length of the streamline that is initialized at
the $y$-coordinate along $S$, and $u_x(y)$ is the $x$ component of the
velocity at the $y$-coordinate along $S$.  Since the flow is laminar in
the absence of grains, the tortuosity of a straight channel is $T=1$.
Following~\cite{dud-koz-mat2011}, two parallel cross-sections are
chosen, and the trajectories that pass between them are formed.  In our
work, we initialize the trajectories at $x=-1$ and terminate them at
$x=1$.

Under the assumptions that the flow is incompressible and not reentrant,
meaning that all streamlines connect the two cross-sections, the
tortuosity can be equivalently measured as~\cite{dud-koz-mat2011}
\begin{align}
  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
           {\displaystyle\int_\Omega u_x(\yy)\, d\yy}.
  \label{eqn:tortuosity2}
\end{align}
Recirculation zones are possible in viscous fluids, but they only
constitute a small fraction of the total porous media, and in our
examples, their velocities are very small.  Therefore, the tortuosity
defined in equation~\eqref{eqn:tortuosity2} is an upper bound for the
definition in equation~\eqref{eqn:tortuosity1}, but for the geometries
we consider, we will show that their difference is negligible (see
Figure~\ref{fig:Eroding20tort_all}).  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories, we use this definition for the
majority of the tortuosity calculations.

The first method to numerically compute the tortuosity uses $N_p$
trajectories that are initially at $(-1,y_i)$, $i=1,\ldots,N_p$, and
stopped when their $x$ coordinate is 1.  We choose $y_i$ so that the
trajectories are initially uniformly spaced in $(-1,1)$.  Having
computed these trajectories, the tortuosity is approximated as
\begin{align}
  T = \frac{1}{d}\frac{\displaystyle\sum_{i=1}^{N_p} 
    u_x(y_i) \lambda(y_i) \Delta y}
  {\displaystyle\sum_{i=1}^{N_p} u_x(y_i) \Delta y_i}, 
\end{align}
where $\Delta y = 2/(N_p + 1)$ and $y_i = -1 + i \Delta y$,
$i=1,\ldots,N_p$.  Alternatively, to compute the tortuosity in terms of
the Eulerian velocity field, we first compute the velocity at $(x_i,y_j)
= (-1 + i\Delta x, -1 + j\Delta y)$, $i,j=1,\ldots,N$, where $\Delta x =
\Delta y = 2/N$, and the velocity at points inside an eroding body are
assigned a value of 0.  Then, the tortuosity is approximated as
\begin{align}
  T = \frac{\sum\limits_{i,j=1}^N \|\uu(x_i,y_j)\|\Delta x \Delta y}
      {\sum\limits_{i,j=1}^N u_x(x_i,y_j) \Delta x \Delta y}.
\end{align}

There have been efforts to relate the tortuosity to the
porosity~\cite{matyka2008tortuosity}, but these results can not be
universal since the porosity which depends on only a single number, the
tortuosity is affected by the distribution of pores within the porous
network.  In addition to the hydraulic tortuosity, a local tortuosity
can be used to analyze the local behavior of a streamline near a
body~\cite{matyka2008tortuosity}.  The local tortuosity of a streamline
initialized at $y$ along $S$ is
\begin{align}
  \tau(y) = \frac{\lambda(y)}{d}.
  \label{eqn:localTort}
\end{align}
\todo[inline]{Need to flesh this out more}
\todo[inline]{Talk about the different relationships between tortuosity
and porosity from the literature}
